---
title: "Theoretical study Quarto -document"
format: pdf
date: "2026-01-28"
editor: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F)

```

### To do 22.2.

-   Add/integrate previous Rmd file of theoretical result as extension
    to this one.

-   Recheck delta t.

-   Check if subindex VAR(1) is necessary for Sigma.

-   Proofs, analysis...

#### Notes:

-   Everything is a scratch and works as a basis for argument, which
    will be clarified when results are in. At this stage result is that
    drift matrix must be symmetric. This is already useful for
    simulations - and very interesting - but we'll see what other things
    turn out. I (Sakari) am not sure at this point still if VAR(1) can
    in facto produce covariance matrix compatible with a common factor
    model, though i consider it very likely.

-   This file will be iterated and grows from autoregression -stuff onto
    the full article. Some foundational layouts are already put in
    place, but should not be treated as set.

### Weak sense stationary VAR(1) and strict longitudinal measurement invariance

In the following VAR(1) process and strict longitudinal measurement
invariance (s-LMI) are compared to each other by the covariance
structure that VAR(1) imposes and how an s-LMI model fits to this (here
equivalently 'is compatible' with) covariance. s-LMI makes sense as a
theoretical model since it captures the simplest scenario of a true
common factor model where only the common factor itself can change.
VAR(1) is also a simple, if not the simplest, vector autoregression
symptom-network model. CT-VAR and VAR can be linked through a
transformation when fixed time intervals are used, or separate study for
CT-VAR can be done.

Especially noteworthy results from mathematical analysis could be
possible constraints that arise and how they can guide empirical
simulations and tell us the reason why s-LMI might not be compatible
with VAR(1) generated data in the first place. Alternatively, if some
VAR(1) models can generate s-LMI compatible data, what constraints are
necessary for the VAR(1) process to produce it?

Add citations to anchor existing research!

Common Factor theoretical model

:   A latent variable $\eta$ is a causal entity which creates observable
    symptoms. Conditional independence of variables
    $P(X_i|\eta,X_j)=P(X_i|\eta):\:i\ne j$ given the common factor
    $\eta$ is the central proposition of the model. All changes in
    symptoms are due to changes in $\eta$. $M_1$ will represent an
    empirical (data-generating) model of the common factor theory.

Mutualistic symptom-network theoretical model

:   Symptoms have causal effects on each others. Symptoms can change by
    themselves or change due to effects not within the symptom-network
    itself. The symptom-network is all the symptoms. $M_2$ will
    represent an empirical (data-generating) model of the
    symptom-network theory.

The models need to be kept vague for now, as the key parts of both
differ in each scenario to be examined. Empirical distinguishability is
here defined - for now - up to the fourth order moment:

Empirical distinguishability

:   $$\begin{aligned} E_{M_1}(X)-E_{M_2}(X)&=D_E\\\   \text{Cov} _{M_1}(X)-\text{Cov} _{M_2}(X)&=D_C  \\   \text{Skew} _{M_1}(X)-\text{Skew} _{M_2}(X)&=D_S\\   \text{Kurt} _{M_1}(X)-\text{Kurt} _{M_2}(X)&=D_S   \end{aligned}  $$

That is, after defining the models, data simulated from both of these
models should not be the same w.r.t. their expectation, covariance,
skewness and kurtosis. Our aim is to identify practical conditions,
where the distinguishability is highest - i.e., when the differences in
either expectation or covariance become largest. We do this in X
different scenarios. In all scenarios we define a respective common
factor theory model as well as a symptom-network mutualism model. We
then aim to analytically show the differences between the models, and
then through simulation quantify what types of practical difference we
might see. We'll be denoting symptoms as random matrix $X$, of which
columns are the symptoms. We use two theoretical models:

The several scenarios relate to study designs, in which both models have
been used using real world empirical data. First, we examine changes
over time. Second, ..., X:th.

### Change across time: Strict Longitudinal Measurement Invariance vs VAR(1) models

Regarding changes in symptoms across time, the most basic models that we
can use for common factor and symptom-network theory are the strict
longitudinal measurement invariance (s-LMI) referred to as and the
vector autoregressive model of order one (VAR(1)). Here we will
analytically approach differences between these models and then perform
simulations to quantify how well empirically the models differ when
generating from a mixture distribution. We will first inspect what type
of covariance structure VAR(1) imposes, and then similarly s-LMI.

Their difference might not always be evident. See for example the
figures 1. and 2. below, where data was generated from both models.

```{r, echo=FALSE}
set.seed(1)
library(mlVAR); library(ggplot2)
X_CF = arima.sim(model = list(ar = sqrt(0.5)), n = 200, n.start = 200, innov = rnorm(n = 200, mean = 0, sd = sqrt(0.5)))
X_CF = matrix(ncol = 3, rnorm(3*200, mean = 0, sd = 1)) + as.numeric(X_CF)

ggplot() + 
  geom_line(aes(y = X_CF[,1], x = seq_along(X_CF[,1]))) +
  geom_line(aes(y = X_CF[,2], x = seq_along(X_CF[,2])), lty = 2) +
  geom_line(aes(y = X_CF[,3], x = seq_along(X_CF[,3])), lty = 3) +
  ylab("X") +
  xlab("Time") +
  ggtitle("1.")

```

In Figure 1, the model was set up so that there is a true common factor
$\eta$ causing three variables $X$. The common factor had an
autoregressive coefficient (of order 1) of $\sqrt{0.5}\approx0.7071$ to
itself and $\psi_t \sim N(0, \sqrt{0.5})$ innovations (0.5 variance).
Factor loadings onto the observed variables were 1. Each of the
$X_{i,t}$ also had their own measurement errors
$\omega_{i,t}\sim N(0,\sqrt{0.5})$. $X_t$ has a covariance matrix
approximately such that the diagonal elements are 2, and off-diagonal
elements are 1.

If simulating from another numerical example, this time a stationary
VAR(1) process, we can observe a fairly similar looking pattern.

```{r, echo = F}
library(fastmatrix)
set.seed(1)
A = matrix(ncol = 4, nrow = 4, 
           c(0.3,0.2,0.2,0.2,
             0.2,0.3,0.2,0.2,
             0.2,0.2,0.3,0.2,
             0.2,0.2,0.2,0.3))
library(mlVAR)
eigen(A)


matrix(solve((diag(1,ncol = 4*4, nrow = 4*4) - kronecker.prod(A))) %*% fastmatrix::vec(diag(c(0.9,0.9,0.9,0.9))),ncol = 4)
A%*%matrix(solve((diag(1,ncol = 4*4, nrow = 4*4) - kronecker.prod(A))) %*% fastmatrix::vec(diag(c(0.9,0.9,0.9,0.9))),ncol = 4)

summary(psych::fa(r = matrix(solve((diag(1,ncol = 4*4, nrow = 4*4) - kronecker.prod(A))) %*% fastmatrix::vec(diag(c(0.9,0.9,0.9,0.9))),ncol = 4), nfactors = 1, n.obs = 10000))

<<<<<<< HEAD
matrix(solve((diag(1,ncol = 3*3, nrow = 3*3) - kronecker.prod(A))) %*% vec(diag(c(0.9,0.9,0.9))),ncol = 3)
=======
>>>>>>> eca457b926e874f202b7b9561fd69703e4ef6ddf

X_VAR = simulateVAR(pars = A, Nt = 200, init = 200, residuals = diag(c(0.9,0.9,0.9)))
ggplot() +
  geom_line(aes(x = seq_along(X_VAR$V1), y = X_VAR$V1)) +
  geom_line(aes(x = seq_along(X_VAR$V1), y = X_VAR$V2), lty = 2) +
  geom_line(aes(x = seq_along(X_VAR$V1), y = X_VAR$V3), lty = 3) +
  ylab("X") +
  xlab("Time") +
  ggtitle("2.")
```

The difference between the two example time series in $D_C$ is

```{r}
D_C = (cov(X_VAR) - cov(X_CF))
dimnames(D_C) = list(paste("X",1:3,sep = ""),paste("X",1:3,sep = ""))
round(D_C, 2)

```

The above is an ensemble covariance, that is, covariance calculated
using all time points simultaneously. This is not the typical case
however. More often we encounter cases where we observe few subsequent
time points and then model them.

# \#### VAR(1) covariance structure

First VAR(1) imposed covariance is derived. Then s-LMI imposed
covariance. Then we move on to inspect how they compare by equating them
together to observe possible contradictions or restrictions. We begin
with such a theoretical view on their imposed covariance structure, and
later move to simulate data from both to see how distinguishable they
are in commonly encountered longitudinal scenarios.

Before the theoretical analysis, a brief substantive consideration. We
will assume (*as Jaakko suggested*) that the symptom-network represented
as the VAR(1) model is a state. By this we mean that it is stationary.
In addition to stationarity, we also assume that there are no external
causes, which would make systematic changes to the symptoms and cause
correlated innovations (or errors, innovations hereon). This assumption
then gives us theoretical circumstance so that the innovations of the
VAR(1) are independent of each other. While this assumption (that no
external events affect the symptom-network) is unlikely to be true in a
real-world scenario especially when considering longer periods of time
(e.g., weeks, months), in any case theoretical analysis would become
near impossible if no restrictions of this kind are made.

#### VAR(1) covariance structure at two subsequent time points

The VAR(1) model is defined in matrix format as
$X_{t}=C+AX_{t-1}+\Gamma_t$, where $\Gamma_t$ is independent error
column vector with $E[\Gamma_t]=0$, $C$ is a constant assumed zero. Also
assume centered $X$, $E[X_t]=0$, in our case. Centering makes covariance
calculations easier as the products of expected values can be mostly
ignored (they become 0). $A$ is $K \times K$ (borrowing from CT-VAR
terminology) 'drift' matrix that includes all lagged effects of
$K\times1$ column vectors $X_t$ to $X_{t-1}$, $K$ being the number of
observed items (symptoms). In this section the focus is on the
$2K\times2K$ covariance matrix where two subsequent measurement time
points are observed. All matrices used are real-valued.

First, the covariance matrix (assumed stationary over time) is

$$
\begin{aligned}
\text{Cov}(X_t) &= E[X_tX_t^T] \\
&= E[(AX_{t-1}+\Gamma_t)(AX_{t-1}+\Gamma_t)^T] \\
&= E[AX_{t-1}X_{t-1}^TA^T] + \underbrace{E[\Gamma_t \Gamma_t^T]}_{=: \Psi} \\
&= AE[X_{t-1}X_{t-1}^T]A^T + \Psi=\Lambda\Lambda^T+\Omega \\
&= \Sigma_t=:\Sigma_{VAR(1)}
\end{aligned}
$$ where stationarity poses that $\Sigma$ is not dependent of time and
so the covariance of VAR(1) is denoted as such from hereon. We will use
$\Sigma_{VAR(1)}$ when we wish to be explicit about meaning the VAR(1)
imposed within time point covariance. $\Gamma_t$ is a random $K\times1$
column vector of (serially) independent innovations at time point $t$,
with $E[\Gamma]=0,\text{Var}(\Gamma=1)$. $\Psi$ is covariance of the
innovations within time point $t$ - i.e., the contemporaneous
covariance. We assumed that the innovations are independent, as
described above, and so $\Psi$ is diagonal.

The vectorized covariance matrix can be solved to equal\
$$
\begin{aligned}
\text{vec}(\Sigma_{VAR(1)})
=
(I-A \otimes A)^{-1} \text{vec}(\Psi)
\end{aligned}
$$

Where vec is the vectorization operator and $\otimes$ is the Kronecker
product. In the above the mixed Kronecker matrix vector product is used
to obtain the result. (Derivation is currently in a scratch file.)

We will equate the VAR(1) posed $\Sigma_{VAR(1)}$ to s-LMI imposed
covariance further down below.

Second, VAR(1) poses that observations at the time points $X_t, X_{t-1}$
have covariance $$
\begin{aligned}
\text{Cov}(X_t,X_{t-1})&=
E[X_tX_{t-1}^T]-E[X_t]E[X_{t-1}]\\&=
E[(AX_{t-1}+\Gamma_t)X_{t-1}^T]\\&=
E[AX_{t-1}X_{t-1}^T]+E[\Gamma_tX_{t-1}^T]\\&=
A\Sigma+E[\Gamma_tX_{t-1}^T]
\end{aligned}
$$ Independent errors means that
$E[\Gamma_tX_{t-1}^T]=\text{Cov}(\Gamma_t,X_{t-1})=0$ leading to $$
\begin{aligned}
\text{Cov}(X_t,X_{t-1})&= A\Sigma_{t-1}\\
\text{Cov}(X_{t-1},X_t)&= \Sigma_{t-1}^T A^T \\
\end{aligned}
$$ Where the two covariances above must be the same - i.e., the
covariance matrix is symmetric. This means that every VAR(1) process
implies that covariance of observations from two subsequent time points
$t, t-1$ is $$
Cov((X_{t-1},X_{t}),(X_{t-1},X_{t})) = 
\begin{pmatrix} 
  \Sigma_{t-1} & A\Sigma_{t-1} \\
  \Sigma_{t-1}^TA^T & \Sigma_t
\end{pmatrix} $$ Where the above covariance matrix is the $2K\times 2K$
covariance matrix of the observed data from the two time points. In
addition, stationarity directly implies $\Sigma_t=\Sigma_{t-1}$. (For
now, notation with sub-index *i* will be kept for clarity as it is.
s-LMI is not necessarily imply stationary, so confusion might be
avoided.)

#### s-LMI covariance structure at two subsequent time points

s-LMI with 1 common factor decomposes $\Sigma_{t-1}$ into following
$$\Sigma=\Lambda\Lambda^T+\Omega_{t-1}$$where by definition of s-LMI
$\Omega_{t-1}$ is diagonal and and $\Lambda$ is a $K\times1$ column
vector of factor loadings constant over time. We also need the
covariance of the common factor at both time points. Let $\delta$ be the
latent regression coefficient which links the common factor to itself at
a previous time point such that $\eta_t=\delta\eta_{t-1}+\psi_t$, where
$\psi_t$ is independent random term ('innovation', 'error',
'disturbance') with $E[\psi_t]=0$. Assuming standardized common factor
such that $E[\eta_{t-1}]=0,\:Var(\eta_{t-1})=1$ covariance of the common
factor at two subsequent time points is

$$
\begin{aligned} 
\text{Cov}(\eta_{t-1},\eta_t)= E[\eta_{t-1}\eta_t]-E[\eta_{t-1}]E[\eta_t]&=\\ E[\eta_{t-1}(\delta\eta_{t-1}+\psi_t)]&=\\ E[\delta\eta_{t-1}^2+\eta_{t-1}\psi_t]&=\\ \delta Var(\eta_{t-1})&= \delta 
\end{aligned}
$$ Now lets look at the $2K\times 2K$ covariance matrix from the
perspective of strict LMI. A s-LMI model imposes that $$
\text{Cov}((X_{t},X_{t-1}),(X_{t},X_{t-1}))= \begin{pmatrix}    \Lambda & 0
\\   0 & \Lambda \end{pmatrix}  \begin{pmatrix}   1 & \delta 
\\   \delta & \delta+Var(\psi_t) \end{pmatrix} \begin{pmatrix}    \Lambda^T & 0
\\   0 & \Lambda^T \end{pmatrix} + \begin{pmatrix}   \Omega_{t-1} & \Omega_{across} 
\\   \Omega_{across}^T & \Omega_t \end{pmatrix} 
$$ where $$
\begin{pmatrix}    \Lambda & 0\\   0 & \Lambda \end{pmatrix}
$$ is a block matrix that sandwiches the $2\times2$ covariance matrix of
the common factor at both time points. $$
\begin{aligned}
&\begin{pmatrix}    \Lambda & 0\\   0 & \Lambda \end{pmatrix} \begin{pmatrix}   1 & \delta \\   \delta & \delta+Var(\psi_t) \end{pmatrix} \begin{pmatrix}    \Lambda^T & 0\\   0 & \Lambda^T \end{pmatrix}+ \begin{pmatrix}   \Omega_{t-1} & \Omega_{across} \\   \Omega_{across}^T & \Omega_t \end{pmatrix} =\\
&\begin{pmatrix}    \Lambda & 0\\   0 & \Lambda \end{pmatrix} \begin{pmatrix}    \Lambda^T & \delta\Lambda^T\\   \Lambda^T\delta & (\delta+Var(\psi_t)\Lambda^T \end{pmatrix} + 
\begin{pmatrix}   
\Omega_{t-1} & \Omega_{across} \\   
\Omega_{across}^T & \Omega_t \end{pmatrix}=\\
& \begin{pmatrix}    
\Lambda\Lambda^T + \Omega_{t-1}& \Lambda\Lambda^T\delta + \Omega_{across}\\   
\Lambda\Lambda^T\delta + \Omega_{across}^T & \Lambda\Lambda^T(\delta+Var(\psi_t)) + \Omega_t 
\end{pmatrix} 
\end{aligned}
$$ From the above we see that the strict LMI can only be compatible with
any process with stationary covariance, if
$\delta+Var(\psi_t)=1\Rightarrow1-\delta=Var(\psi_t)$ (assuming
$\Lambda$ is non-zero). (When fitting a s-LMI model this is allowed.) We
also see that s-LMI is compatible with non-stationary processes where
the covariance is proportional to $\delta+Var(\psi_t)$ aligning with
previous theoretical analysis where covariance increased over time in a
LMI preserving model.

A brief note on notation: We'll be using simply $\Omega$ for the s-LMI
residual covariance, since residual covariance is assumed invariant over
time $\Omega_{t-1}=\Omega_{t}=\Omega_{t+1}=…=\Omega$ .

Using the above auxiliary results we can move to analyse the null
hypothesis (hypotheses) of no difference between VAR(1) and s-LMI.

Working null hypothesis (1)

:   If covariance at some (measurement) time point is perfectly
    explained by a common factor model with non-zero factor loadings
    $\Lambda$, and if the data generation is from a stationary VAR(1)
    model, then s-LMI model fits perfectly.

Considering only the subset of VAR(1) processes which create a
covariance matrix that can be perfectly explained by a common factor
model is done because we're interested in how (if at all) VAR(1) can
deviate from s-LMI in terms of produced data. Understandably, if any
VAR(1) model creates covariance structure incompatible with s-LMI model
at some time point (i.e., a covariance matrix non-compatible with a
common factor model), then deviation must occur (although the extent to
which this occurs is not clear at this point).

If the VAR(1) generated $2K\times2K$ matrix cannot be explained by the
strict LMI model, this seems likely to be because the off diagonal
blocks of covariance matrices across time points are non-compatible with
the respective s-LMI model imposed across time covariance. Combined with
the restriction on the within time point covariance, this might lead to
contradictions.

This gives us the following null hypothesis (1) equations (from the
$2K\times2K$ matrices imposed by VAR(1) and s-LMI)

$$
\begin{aligned}
(I-A \otimes A)^{-1} \text{vec}(\Psi) &= \text{vec}(\Lambda \Lambda^T + \Omega)&&\Rightarrow
\\
(I-A \otimes A)^{-1} \text{vec}(\Psi) &= \text{vec}(\Lambda \Lambda^T) + \text{vec}(\Omega)
\\
AE[X_{t-1}X_{t-1}^T]A^T + \Psi&=\Lambda \Lambda^T + \Omega\Rightarrow
\\
A\Sigma A^T+\Psi&=\Sigma\Rightarrow\\
(A\otimes A)\text{vec}(\Sigma)&=\text{vec}(\Sigma)-\text{vec}(\Psi)
\end{aligned}
$$ and $$
\begin{aligned}
&A\Sigma_{t-1}=\Lambda\Lambda^T\delta+\Omega_{across}
\end{aligned}
$$ both of which must be true for the null hypothesis (1) to hold.
Assuming that the null hypothesis (1) is true, further analysis of the
respective equations show $$
\begin{aligned}
A\Sigma_{t-1} &= \Lambda\Lambda^T\delta+\Omega_{cross} \Leftrightarrow\\
A &= \Lambda\Lambda^T\delta \Sigma_{t-1}^{-1} + \Omega_{cross} \Sigma_{t-1}^{-1} \Leftrightarrow\\
A + \delta \Omega \Sigma_{t-1}^{-1} &= \Lambda\Lambda^T\delta \Sigma_{t-1}^{-1} + \delta \Omega \Sigma_{t-1}^{-1} + \Omega_{cross} \Sigma_{t-1}^{-1} \Leftrightarrow\\
A + \delta \Omega \Sigma_{t-1}^{-1} &= \delta \underbrace{(\Lambda\Lambda^T + \Omega)}_{=\Sigma_{t-1} \text{ by assumption}} \Sigma_{t-1}^{-1} + \Omega_{cross} \Sigma_{t-1}^{-1} \Leftrightarrow\\
A &= \delta I + (\Omega_{cross} - \delta \Omega)\Sigma_{t-1}^{-1}.
\end{aligned}
$$

We'll see if contradictions arise when including three time points in
the analysis below. But first, we need to do some generalizations and
discuss the implications of VAR(1) model and the null hypothesis (1)
further.

#### VAR(1) covariance compared to s-LMI covariance at T time points.

As the scenario where two subsequent measurement points are observed is
possibly the most common one, we'll keep the above discussion in place
for now. On the other hand, it is of interest to analyze what happens
when multiple subsequent time points are included. This is perhaps less
common in measurement invariance literature, but more common in VAR
literature.

We have (proven below) that as the distance in time between two time
points $\Delta t=(t_i-t_j)\to\infty, \:i>j$ samples from VAR(1) model
generated data at those two time points have 0 covariance. This would
make the asymptotic $2K\times2K$ perfectly explained by a measurement
invariance model, because the main diagonal covariance matrices
$\Sigma_{t-1}=\Sigma_t$ are perfectly explained by a common factor model
by assumption, and the off diagonal matrices are 0, which is allowed in
a strict LMI model (no cross-covariance between the observed time points
and 0 regression coefficient for the latent variable). Again, we are
assuming that VAR(1) model generated $\Sigma$ perfectly compatible with
s-LMI. This is a less practically meaningful case arguably since
observed data with 0 across time point covariance is not common. This
also does prove that a VAR(1) process could in some sense be the true
data generating process even if no across time point covariance is
observed (no lagged effects are estimated) if one were to claim that
$\Delta t$ is very small: We're just not observing time points close
enough to each other to see the VAR(1).

Nevertheless - considering that the above scenario is not a typical one
in psychopathology research - we can attempt to generalize the above
results concerning the $2K\times2K$ matrix to an $TK\times TK$ matrix
where we have $T$ measurement of $K$ symptoms over occasions at constant
time intervals. Brief notes are made as we move on and a summary at the
end.

Here we change the notation a little and use an arbitrary time point $t$
as the 'first' measurement time point, so that time points increase
${t, t+1,t+2,...,T}$, where $T$ is the last measurement time point. The
$TK\times TK$ matrix is

$$
\begin{pmatrix}
  \Sigma_t & \dots & \Sigma_{t,T}
  \\ \vdots&\ddots&\vdots
  \\ \Sigma_{t,T}^T&\dots&\Sigma_T 
\end{pmatrix}
$$

Proceeding again from VAR(1) to s-LMI and then equating between the
models. Let $^{(c)}$ denote the matrix raised to power of $c$ and
$\Delta t=t_k-t_j,\:j<k$. The $\Sigma_{t,T}$ for VAR(1) is

$$
\Sigma_{t,T}=\text{Cov}(X_j, X_{k}) = A^{(\Delta t)}\Sigma_{VAR(1)}
$$

*Proof of the asymptotic 0 across time point covariance*. On a brief
note we can further decompose the above equation, using the power method
of eigenvalues, into

$$
PD^{(T)}P^{-1}\Sigma_{VAR(1)}
$$

where $D$ is a diagonal matrix of eigenvalues of $A$, $P$ is and
orthonormal matrix of eigenvectors of $A$ as columns. From this
decomposition we directly see the above mentioned asymptotic property
that as distance in time between time points increases, $D$ is raised to
a larger power and decreases eventually to the zero matrix. This is
because eigenvalues of (stationary) $A$ are less than one, meaning that
all diagonal elements of $\text{diag}(D):|d_{ii}|<1$ as well, and hence
the matrix power converges to $0$. This means that a true data
generating stationary VAR(1) model should produce across time point
covariances that reduce to zero as distance in time between any two time
points increases (practically speaking perhaps as years pass). Such
observations might be sparse in psychopathology literature and it is
accepted in the literature that even if any autoregressive model would
be the true data generating model, it would unlikely be stationary over
lengthier time periods (e.g., years again). This property of stationary
VAR(1) does - as we have - restrain the model to analysis of a
psychopathological state, but this contemplation is omitted here.

For s-LMI, respectively, we'll use
$\delta_{2}, \delta_{3}, ..., \delta_t,\delta_{t+1},...,\delta_{T}$ to
denote the regression coefficient between subsequent time points (there
is no regression at the first time point in s-LMI). Also, the
$\Omega_{t,t+1}$ needs to be generalized to include residual covariances
between any two time points so that $\Omega_{t,T}$ is the residual
covariance between time point $t$ and time point $T$. $\Omega$ is the
within time point residual covariance invariant over time.

Few more generalizations and constraints before we can equate the VAR(1)
implied covariance to the s-LMI implied covariance again using multiple
time points. We have that the VAR(1) imposed covariance between any two
subsequent time points is the same. This also generalizes to the VAR(1)
imposed covariance between any two equidistant time points as they are
$A^{\Delta t}\Sigma$, which only depends on the distance in time. This
means that $\delta_t$ must be a constant since otherwise the s-LMI
imposed covariance between two subsequent time points
$\Lambda\Lambda^T\delta + \Omega_{t,t+1}$ would not be the same. More
precisely, $\Omega_{t,t+1}$ is by definition diagonal and so can only
change the diagonal to some extent - off-diagonal elements would not be
the same. This also means that $\Omega_{t,t+1}$ is the same for all $t$.

Using that $\delta$ is constant (and other assumptions established
above), the covariance between the common factor to itself between any
time points two time intervals apart from each other is

$$
\begin{aligned}
\text{Cov}(\eta_t,\eta_{t+2})&=E[\eta_{t}\eta_{t+2}]\\
&=E[\eta_{t}  (\delta_{t+2}\eta_{t+1}+\psi_{t+2})  ]\\
&=E[\eta_{t}  (\delta_{t+2}(\delta_{t+1}\eta_{t}+\psi_{t+1})+\psi_{t+2})  ]\\
&=E[\eta_{t}  (\delta_{t+2}\delta_{t+1}\eta_{t}+\delta_{t+2}\psi_{t+1}+\psi_{t+2})  ]\\
&=E[\delta_{t+2}\delta_{t+1}\eta_{t}\eta_{t}]+E[\delta_{t+2}\psi_{t+1}\eta_{t}]+E[\psi_{t+2}\eta_{t}]\\
&=\delta_{t+2}\delta_{t+1}=\delta^2
\end{aligned}
$$

and for three time intervals apart

$$
\begin{aligned}
\text{Cov}(\eta_t,\eta_{t+3})&=E[\eta_{t}\eta_{t+3}]\\
&=E[\eta_{t}  (\delta\eta_{t+2}+\psi_{t+3})  ]\\
&=E[\eta_{t}  (\delta(\delta\eta_{t+1}+\psi_{t+2})+\psi_{t+3})  ]\\
&=E[\eta_{t}  (\delta(\delta(\delta\eta_{t}+\psi_t)+\psi_{t+2})+\psi_{t+3})  ]\\
&=E[\eta_{t}  (\delta(\delta^2\eta_{t}+\delta\psi_t)+\psi_{t+2})+\psi_{t+3})  ]\\
&=E[\eta_{t}  (\delta^3\eta_{t}+\delta^2\psi_t+\delta\psi_{t+2}+\psi_{t+3})  ]\\
&=E[\eta_{t}\delta^3\eta_{t}+\eta_{t}\delta^2\psi_t+\eta_{t}\delta\psi_{t+2}+\eta_{t}\psi_{t+3})  ]\\
&=\delta^3=\text{Cov}(\eta_t,\eta_{t+2})\delta
\end{aligned}
$$

which can be by induction (not currently properly done) shown to result
in

$$
\begin{aligned}
\text{Cov}(\eta_t,\eta_{t+\Delta t})&=\delta^{\Delta t}
\end{aligned}
$$

This implies that the covariance between any two time points must be
$\Lambda\Lambda^T\delta^{\Delta t} + \Omega_{t,T}$.

From the previous result for the $2K\times2K$ matrix we can then
generalize to the $TK\times TK$ matrix

$$
\begin{pmatrix}
   \Lambda\Lambda^T + \Omega & ... &  \Lambda\Lambda^T \delta^{(T-1)} + \Omega_{1,T}
\\ \vdots&\ddots&\vdots
\\ \Lambda\Lambda^T \delta^{(T-1)} + \Omega_{1,T}^T&...&\Lambda\Lambda^T(\delta+Var(\psi_t)) + \Omega
\end{pmatrix}
$$

Now we can obtain the more general, null hypothesis equation relating
the across time point covariances and within time point covariances for
any lag

$$
\begin{aligned}
\Sigma_{\Delta t}&=\\
A^{(\Delta t)}\Sigma&=\Lambda\Lambda^T \delta^{\Delta t}+ \Omega_{\Delta t}&&
\end{aligned}
$$

This directly tells us that $A$ must be symmetric, positive definite:

$$
\begin{aligned}
            & A^{(\Delta t)}\Sigma&=\Lambda\Lambda^T \delta^{\Delta t}+ \Omega_{\Delta t}&&\\
\Rightarrow & A^{(\Delta t)}&=(\Lambda\Lambda^T \delta^{\Delta t}+ \Omega_{\Delta t})\Sigma^{-1}&&\\
\end{aligned}
$$

where the right hand side, if $\delta$ is positive, is a multiplication
of two positive-definite (covariance) symmetric square matrices, which
must also be positive definite and symmetric.

#### Discussion notes

The above result has the substantive implications, that a
symptom-network VAR(1) model can only be compatible with a common factor
model in the restricted case of all symptoms affecting each other with
bi-directional effects which are equivalently strong in each direction
$X_i\to X_j=X_i\leftarrow X_j$. That is, distinguishability $D_C$ will
be larger the more asymmetric the drift matrix $A$ is.

Looking from a measurement perspective, we might be interested what
happens in a 0 measurement error scenario. So assume there is no unique
variance in $X$. In this case we obtain

$$
\begin{aligned}
&A^{(\Delta t)}\Sigma=\Lambda\Lambda^T \delta^{\Delta t}+ \underbrace{\Omega_{\Delta t}}_{=0}&\\     
\Rightarrow&A^{(\Delta t)}\Sigma=\underbrace {\Lambda\Lambda^T}_{\text{By assumption: }\Sigma-\Omega=\Sigma} \delta^{\Delta t}&\\
\Rightarrow&A^{(\Delta t)}\Sigma=\delta^{\Delta t}\Sigma&\\
\Rightarrow&A^{(\Delta t)}=\delta^{\Delta t}I&
\end{aligned}
$$

that is, $A$ must be a scalar multiple of the identity matrix. This
would lead to a contradiction (if we assume that there must be positive
covariance between symptoms at least), since now there should be 0
covariances between symptoms.

#### Without measurement error

WE SHOULD ASSUME NO SERIAL COVARIANCE OF ERRORS! If we look at a theoretical scenario where we observe psychopathology without measurement error, and
assume that there is no unique variance ($\Omega=0$), then $$
\begin{aligned}
&A^{(\Delta t)}\Sigma=\Lambda\Lambda^T \delta^{\Delta t}+ \underbrace{\Omega_{\Delta t}}_{=0}&\\     
\Rightarrow&A^{(\Delta t)}\Sigma=\underbrace {\Lambda\Lambda^T}_{\text{By assumption: }\Sigma-\Omega=\Sigma} \delta^{\Delta t}&\\
\Rightarrow&A^{(\Delta t)}\Sigma=\delta^{\Delta t}\Sigma&\\
\end{aligned}
$$ 
We see that $A$ must hence be scalar multiple of the identity matrix
(or a scalar). Since $\Sigma =\Lambda\Lambda^T$ this means, that $\Lambda$ must be an eigenvector of $A$.

This is a contradiction, because a diagonal $A$ and
diagonal $\Psi$ make covariances between the variables zero. This means
that $\Lambda$ must also be zero. This is against our assumption of null
hypothesis (1).

If measurement error exists ($\Omega\ne0$)

$$
\begin{aligned}
&A^{(\Delta t)}\Sigma=\Lambda\Lambda^T \delta^{\Delta t}&\\     
\end{aligned}
$$

where we observe that the left hand side is rank 1, since
$\Lambda\Lambda^T$ is a product of two column vectors.
$\Lambda\Lambda^T$ has one non-zero eigenvalue (the magnitude of the
common factor).

Rewriting and differentiating w.r.t. $\Sigma$ we obtain that

$$
\begin{aligned}
&A^{(\Delta t)}\Sigma=\Lambda\Lambda^T \delta^{\Delta t}+ \Omega_{\Delta t}&\\  
\Rightarrow& A^{(\Delta t)} \Sigma = (\Sigma-\Omega) \delta^{\Delta t}+ \Omega_{\Delta t}&\\  
\Rightarrow& A^{(\Delta t)} \Sigma = (\Sigma-\Omega) \delta^{\Delta t}+ \Omega_{\Delta t}&\\  
\Rightarrow& A^{(\Delta t)} \Sigma = \Sigma \delta^{\Delta t} -\Omega \delta^{\Delta t}+ \Omega_{\Delta t}&(\frac{\partial}{\partial\Sigma})\\  
\Rightarrow& A^{(\Delta t)}  = \delta^{\Delta t} \\  
\end{aligned}
$$ We observe that $A$ must be diagonal, and this leads again to a
contradiction.

### Summary and implications OLD 31.1.2024

-   A direct observation from the null hypothesis (1) and auxiliary null
    hypothesis (A1) equations shown above is that $A$ must always be
    symmetric for the common factor s-LMI model to be compatible with
    stationary VAR(1) generated data. Hence, asymmetry in $A$ produces
    incompatibility to s-LMI.

-   s-LMI is compatible with the idea that VAR(1) poses that across time
    point covariance approaches 0 as the distance between two time
    points increases. This scenario might not be frequently observed
    however, suggesting that - at least some part of - psychopathology
    processes cannot be understood as vector autoregressive processes.
    This is true for any stationary process to my knowledge, as the
    requirement is that the process will necessarily remain bounded to
    some extent to its 'state'. In fact, it might be reasonable to
    integrate both viewpoints as is done in trait-state models.

### Next

-   Perhaps one possibility is something like Chi-squared testing with
    vectorized (non-redundant) squared elements of
    $A_{lower-tri}^T-A_{upper-tri}$.
-   At this point it still is not evident that VAR(1) can generate a
    covariance matrix perfectly decomposable to a common factor
    model(?). Even more, if this is possible for multiple subsequent
    timepoints is unclear.
-   Also it is not clear what happens when sampling from a population at
    different time points so that subjects are sampled at different time
    lags.
-   Stationarity is also not necessarily a condition which should be
    imposed, which can be discussed further.
-   VAR(1) is a special case of CT-VAR. The respective transformations
    from CT-VAR to VAR(1) are available.
-   Measures of asymmetry such as $s$ $$
    s \equiv (|A_{sym}|-|A_{anti}|)/(|A_{sym}|+|A_{anti}|)
    $$ {#eq-asymmetry} where $A$ is decomposed into its symmetric and
    asymmetric parts can be used also. The metric above is shown at:
    [stack
    exchange](https://math.stackexchange.com/questions/2048817/metric-for-how-symmetric-a-matrix-is).
    Asymmetry of $A$ could further be approached analytically for
    example by decomposing $A$ into its symmetric and asymmetric parts,
    and/or through simulations where asymmetry of $A$ is varied by
    producing random matrices with some logic with how asymmetric is
    produced in $A$.
-   

#### 
